---
title: "EDF 5484, 9.3"
format: html
editor: visual
---

9.3 Uncertainty in the predicted expectation and the forecast: Consider the economy and voting example from Section 7.1. Fit the linear regression model to the data through 2012; these are available in the folder ElectionsEconomy.

```{r}

hibbs <- read.table("https://raw.githubusercontent.com/avehtari/ROS-Examples/master/ElectionsEconomy/data/hibbs.dat", header=TRUE)

View(hibbs)

library(rstanarm)
M1 <- stan_glm(vote ~ growth, data=hibbs)
print(M1)

```

(a) Make a forecast for the incumbent party’s share of the two-party vote in a future election where economic growth is 2%.

```{r}
new <- data.frame(growth=2.0)
y_point_pred <- predict(M1, newdata=new)
y_point_pred # predicted outcome is 52.38

# If we calculate manually.... 
a_hat <- coef(M1)[1]
b_hat <- coef(M1)[2]
y_point_pred <- a_hat + b_hat*2
y_point_pred
#We get thew same results
```

(b) Compute the point forecast, the standard deviation of the predicted expectation from (9.1), and the standard deviation of the predicted value from (9.2).

```{r}
pred <- predict(M1,newdata=new,se.fit=TRUE)
sigma <- M1$stan_summary["sigma","mean"]
## Expected Value Prediction Error
pred$fit + qt(c(.05,.95),nrow(hibbs)-2)*pred$se.fit
## Individual Prediction Year
pred$fit + qt(c(.05,.95),nrow(hibbs)-2)*
  sqrt(pred$se.fit^2+sigma^2)


```

(c) Now compute these using the relevant prediction functions discussed in Section 9.2. Check that you get the same values as in part (a) of this problem.

```{r}

```
