---
title: "Matrixes"
format: html
---

## is.na and anyNA

```{r}
mmat <- matrix(1:12,3,4)
mmat[2,3] <-NA
mmat
is.na(mmat) ## Individual elements
anyNA(mmat) ## Whole data set.
```

Don't use `if (is.na(X)) {...}` use `if(anyNA(X)) {...}` instead.

## NAs are contageos

```{r}
mmat+1
mmat >5

mmat == NA
0/0
log(0)
```

```{r}
mean(mmat)
mean(mmat,na.rm=TRUE)
rowSums(mmat)
rowSums(mmat,na.rm=TRUE)
```

is.na(mmat[1,1]) <- TRUE
mmat
is.na(mmat[1,1]) <- FALSE ## This doesn't work
mmat[1,1] <- 1
mmat

## Making matrixes

```{r}
matrix(1:12,3,4)
matrix(1:12,3,4,byrow=TRUE)
```
```{r}
Nrows <- 100
Ncols <- 5
rmat <- matrix(rnorm(Nrows*Ncols),Nrows,Ncols)
colnames(rmat) <- paste("var",1:Ncols,sep="")
head(rmat)
summary(rmat)
```

## Data frames
```{r}
df <- data.frame(x=1:50, y=rnorm(50),
                 z=runif(50)<.5)
head(df)
head(df$y)
```

For a matrix, you need to use [], not $
```{r}
rmat[1:6,"var2"]
```



```{r}
df$w <- NA
head(df)
head(df[1])
head(df[c(1,3)])
df[1,3]
```
## Rep
rep(1:3,2)
rep(1:3,each=2)
rep(c("control","experimental"),each=8)
arm <- factor(rep(c("control","experimental"),each=8))


hml <- factor(c("high","high","med",
                   "med","low","low"))
as.numeric(hml)
hml <- ordered(c("high","high","med",
                   "med","low","low"),
                  levels=c("low","med","high"))
as.numeric(hml)


## apply
apply(mmat,1,sum) ## row sums
apply(mmat,2,sum) ## column sums
mmat
apply(mmat,1,sum,na.rm=TRUE) ## row sums
apply(mmat,2,sum,na.rm=TRUE) ## column sums

sapply(df,mean,na.rm=TRUE)

as.data.frame(mmat)
as.matrix(df)
