---
title:  Case Study 3 -- Regression Diagnostics
format:
  pdf: 
    header-includes:
      - \usepackage{annotate-equations}
---

```{r}
library(tidyverse)
library(DescTools)
library(plotly)
```


## Prediction Error

$$
s_{\hat{y}_p} = \sqrt{
{s^2}
\left [
{\frac{1}{n}} +
{\frac{(x_p - \bar{X})^2}{SS_{X}}} +
{1}
\right ]
}
$$
## Car stopping distance

Three models
```{r}
cars.lm <- lm(dist~speed,data=cars)
cars.sqm <- lm(sqrt(dist)~speed,data=cars)
cars.llm <- lm(log(dist)~speed,data=cars)
```

### Linear

```{r}
ggplot(cars,aes(x=speed,y=dist)) + geom_point() + geom_smooth()
```
If you have loaded the plotly package, you can get an interactive version of the last scatterplot with `ggplotly()` [I've turned it off here to generate pdf output, only works for html output.]

```{r plotly, eval=FALSE}
ggplotly()
```


```{r}
plot(cars.lm)
```

## Square Root

```{r}
ggplot(cars,aes(x=speed,y=sqrt(dist))) + geom_point() + geom_smooth()
```


```{r}
plot(cars.sqm)
```

## Log

```{r}
ggplot(cars,aes(x=speed,y=log(dist))) + geom_point() + geom_smooth()
```


```{r}
plot(cars.llm)
```

# ACED Data

```{r loadACEDnoNA}
ACEDextract <- read_csv("ACED_extract1.csv",na="-999")
ACEDextract$Session <- factor(ACEDextract$Session)
ACEDextract$Cond_code <- factor(ACEDextract$Cond_code)
ACEDextract$Sequencing <- factor(ACEDextract$Sequencing)
ACEDextract$Feedback <- factor(ACEDextract$Feedback)
ACEDextract$Gender <- factor(ACEDextract$Gender)
ACEDextract$Race <- factor(ACEDextract$Race,1:8)
ACEDextract$Level_Code <- factor(ACEDextract$Level_Code)
```

```{r mutate}
ACEDextract %>%
  mutate(gain=post_scaled-pre_scaled) ->
  ACEDextract
```


# dfbetas




```{r dfbetas}
lm_PostxEAP <- lm(EAP.sgp~post_scaled,data=ACEDextract,na.action=na.exclude)
dfb <- dfbeta(lm_PostxEAP)
dfbs <- dfbetas(lm_PostxEAP)
data.frame(dfbeta=dfb,dfbeta_scaled=dfbs) ->dfbb
summary(dfbb)
dfbb
```

```{r}
boxplot(dfb[,"post_scaled"])
```
```{r}
leverage <- cut(dfb[,"post_scaled"],quantile(dfb[,"post_scaled"],probs=c(0,.05,.95,1),na.rm=TRUE))
summary(leverage)
ggplot(ACEDextract,aes(y=pre_scaled,x=EAP.sgp,color=leverage)) + scale_color_discrete() + geom_point() + geom_smooth(method="lm") 
```
```{r plotLm}
plot(lm_PostxEAP)
```


# Five Assumptions

## 0.  Representative Sample

### Missing Data

## 1. Independent Observations

### Time Series

### Clusters

## 2. Linear form

### Transformations

## 3. Homoscedasticiy -- Equal variances

### Weighted regression

### Transforming Y

## 4. Normality of Residuals

## 5. Subgroup differences

### Color plots by group

Do the groups mix?  

Are the slopes similar?

### Analysis of Covariance (look ahead)

# Outliers and influential points.

## Sensisitivity Analysis

# Assignment

For the regression `post_scaled ~ EAP.sgp`; answer the following questions?

0. Are there problems with the sample?  Non-representative group?  Missing data? 

Sample is a single middle school in a wealthy NJ district, so low generalizability to all middle schools.

Most missing data is because control students did not get internal ACED measure (EAP.sgp)

1. Is the distribution the same for everybody in the group?  Are there clusters?  Serial (time) dependencies?

Grouped by classrooms.  (Might want to color by class.)

```{r groupingGender}
ggplot(ACEDextract,aes(x=EAP.sgp,y=post_scaled,color=Gender)) + geom_point()+geom_smooth()
```

```{r groupingClass}
ggplot(ACEDextract,aes(x=EAP.sgp,y=post_scaled,color=Session)) + geom_point()
```

```{r ecologicalCorr}
class1 <- data.frame(x=rnorm(25,5),y=rnorm(25,5),class="c1")
class2 <- data.frame(x=rnorm(25,10),y=rnorm(25,10),class="c2")
bothclasses <- rbind(class1,class2)
bothclasses$class <- factor(bothclasses$class)
ggplot(bothclasses, aes(x=x,y=y,color=class)) + geom_point()
```




2. Is the relationship mostly linear?  Will transforming X or Y make it more linear?

3. Is the variance roughly the same for all values of $\hat Y$?  Will transforming $Y$ help?

4. Are the residuals roughly normal?  Are there any outliers?  Will transforming $Y$ help?

5. Are there any differences by subgroup?  

In the ACED data set, pay attention to `Level_Code` and `Cond_code`.

```{r groupingCondition}
ggplot(ACEDextract,aes(x=EAP.sgp,y=post_scaled,color=Cond_code)) + geom_point()+geom_smooth()
```
